---
title: "Final Assessment PHY 125.3"
author: "Prasham Bhuta"
output:
  pdf_document: default
  html_notebook: default
---
# Final Assessment - Probability
This is the final assessment to the probability chapter of Data Science course
at Harvardx.

Loading necessary libraries.
```{r library, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(dslabs)
```

Loading the '2015 US Period Life Table `(death_prob)`' from dslabs and 
summarising it.

```{r load_data}
data("death_prob")
head(death_prob)
summary(death_prob)
```
## The scenario

Insurance company offers a 1 year term policy that pays $150000 in event of death. The premium for a 50 year old female is $1,150 yearly. 
In event of a claim, the company forfeits the premium, and pays $150000 as agreed. The company plans to sell 1000 such insurances.

Lets setup the basic parameters.

```{r setting_up}
n <- 1000
loss_payout <- -150000
profit_premium <- 1150
```

Lets plot age vs probability of death.
```{r plot1}
death_prob %>% ggplot(aes(age, prob, group = sex, color = sex)) + 
    geom_line() +
    ggtitle('Age vs Probability of death')
```

## Q1(a): Use `death_prob` to determine the death probability of 50 year old female.
```{r 1a}
death_prob %>% filter(age == 50 & sex =='Female')
```
## Q1(b): What is the Expected Value(E) for a 50 year old female
```{r 1b}
p <- 0.003193
loss_payout*p + profit_premium*(1-p)
```

## Q1(c): Calculate the Standard Error (SE) for the same.
```{r 1c}
sqrt(p*(1-p))*abs(loss_payout - profit_premium)
```

## Q1(d): (E) for all 1000 policies for 50 year old females
```{r 1d}
En <- n*(loss_payout*p + profit_premium*(1-p))
En
```

## Q1(e): (SE) for all 1000 policies for 50 year old females.
```{r 1e}
SEn <- sqrt(n*p*(1-p))*abs(loss_payout - profit_premium)
SEn
```

## Q1(f): Using Central Limit Theorem to calculate the probability of losing money on this set of 1000 policies.
```{r 1f}
pnorm(0, En, SEn)
```

## Section 2: Lets perform similar tasks for 50 year old Male. 

## Q2(a): Probability of death for a 50 year old Male
```{r 2a}
prob <- death_prob %>% filter(age == 50 & sex == 'Male')
prob
```
## Q2(b): Suppose from 1000 policies the expected profit is to be $700000. What should be the premium?
```{r 2b}
# En = n * (loss*p + premium*(1-p))
# premium = (En - n*loss*p)/(n*(1-p))
En <- 700000
p <- prob$prob
premium <- (700000 - n*loss_payout*p)/(n*(1 - p))
premium
```
## Q2(c): calculate the SE for n policies with previous premium amount.
```{r 2c}
SEn <- sqrt(n*p*(1-p))*abs(loss_payout-premium)
SEn
```
## Q2(d): Using Central Limit Theorem what is the probability of losing money on 1000 policies.
```{r 2d}
pnorm(0, En, SEn)
```

## Section 3: An event changes the death_prob, what will be change in profit for the insurance company.
```{r section3}
new_p <- 0.015
loss_payout <- -150000
premium <- 1150
n <- 1000
```
## Q3(a): What is the new expected value?
```{r 3a}
# E = n*(loss*new_p + premium*(1-new_p))
En <- n*(loss_payout*new_p + premium*(1-new_p))
En
```
## Q3(b): What is the new expected value?
```{r 3b}
# SE = sqrt(n*new_p*(1-new_p))*abs(loss-premium)
SEn <- sqrt(n*new_p*(1-new_p))*abs(loss_payout-premium)
SEn
```

## Q3(c): What is the probability of company losing money.
```{r 3c}
pnorm(0, En, SEn)
```

## Q3(d): What is the probability of losing more than 1 million $?
```{r 3d}
pnorm(-1000000, En, SEn)
```

## Q3(e): New death probability are as defined below, what is the lowest death probability for which chance of losing money exceed 90%?
```{r 3e}
p <- seq(.01, .03, .001)
En <- n*(p*loss_payout + premium*(1-p))
SEn <- sqrt(n*p*(1-p))*abs(loss_payout - premium)
prob_losing_money <- pnorm(0, En, SEn)
df <- data.frame(prob = p, losing_money = prob_losing_money)
df %>% filter(losing_money>0.9)
```
## Q3(f): For the new death probabilities what is the least death probability for which the chance to lose over 1million$ exceeds 90%?
```{r 3f}
p <- seq(.01, .03, .0025)
En <- n*(p*loss_payout + premium*(1-p))
SEn <- sqrt(n*p*(1-p))*abs(loss_payout - premium)
prob_losing_money <- pnorm(-1000000, En, SEn)
df <- data.frame(prob = p, losing_money = prob_losing_money)
df %>% filter(losing_money>0.9)
```

## Q4(a): Define a sampling model for n = 1000 loans, p_loss = 0.015, loss = -150000$, premium = 1150$. What is the reported profit in millions?

Set the seed to 25 using following command:
set.seed(25, sample.kind = "Rounding").

```{r, warning=FALSE}
set.seed(25, sample.kind = "Rounding")
```

```{r 4a}
n <- 1000
p_loss <- 0.015
loss <- 150000
premium <- 1150
report <- sample(c(-1, 1), n, replace = TRUE, 
                          prob = c(p_loss, 1-p_loss))
profit <- 1150*sum(report == 1)/10^6
loss <- 150000*sum(report == -1)/10^6
profit - loss
```

## Q4(b): set the seed to 27, and run the simulation with 10000 replicates. Finally, find out what is the probability of losing 1 million $ or more?
```{r}
set.seed(27, sample.kind = 'Rounding')
```


```{r 4b}
B <- 10000
profits <- replicate(B, {
    report <- sample(c(-1, 1), n, replace = TRUE, 
                     prob = c(p_loss, 1 - p_loss))
    profit <- 1150*sum(report == 1)/10^6
    loss <- 150000*sum(report == -1)/10^6
    total <- profit - loss
})
mean(profits <= -1)
```
## Q5(a): For a death prob of 0.015, the probability of losing money should be less than 5%. Calculate the premium required.

Pr(S < 0) = 0.05

Pr((S - En)/SEn < -En/SEn) = 0.05

Pr(Z < - En/SEn) = 0.05

therefore, -En/SEn = qnorm(0.05)

```{r 5a}
# -[(loss*p + premium*(1-p))*n/(sqrt(n*p*(1-p))*abs(loss-premium)) = qnorm(0.05)
loss_payout <- -150000
p <- 0.015
n <- 1000
z <- qnorm(0.05)

# -(-2250000 + 985*premium)/(3.843*premium + 576570) = -1.6448
# 2250000 - 985*premium = -6.320966*premium - 948342
# premium = (2250000 + 948342)/(985 - 6.320966)
premium <- (2250000 + 948342)/(985 - 6.30966)
premium
```
## Q5(b): Expected profit per policy at above premium rate?
```{r 5b}
En <- loss_payout*p + premium*(1-p)
En
```
## Q5(c): What is the expected profit for over 1000 policies?
```{r 5c}
En*1000
```
## Q5(d): Using Monte Carlo Simulation determine the probability of losing money for 10000 trials.
```{r}
set.seed(28, sample.kind = "Rounding")
```
```{r 5d}
B <- 10000
n <- 1000
loss_payout <- -150000
premium <- 3267.982
p <- 0.015

profits <- replicate(B, {
    draws <- sample(c(loss_payout, premium), n, replace = TRUE, 
                    prob = c(p, 1-p))
    sum(draws)
})
mean(profits < 0)
```

## Section 6: The pandemic changes the probability of payout by a value between -0.01 to 0.01, using this changes calculate the profit returns.

Set seed to 29.

```{r}
set.seed(29, sample.kind = "Rounding")
```
## Q6(a): What is the expected value for 1000 policies?

```{r 6a}
B <- 10000
n <- 1000
loss_payout <- -150000
premium <- 3267.982

profits <- replicate(B, {
    p <- 0.015 + sample(seq(-0.01, 0.01, length = 100), 1)
    draws <- sample(c(loss_payout, premium), n, replace = TRUE,
                    prob = c(p, 1-p))
    sum(draws)
})
mean(profits)
```

## Q6(b): What is the probability of losing money?
```{r 6b}
mean(profits < 0)
```

## Q6(c): What is the probability of losing more than 1 million $?
```{r 6c}
mean(profits < -1000000)
```
