---
title: "US-murders ggplot"
author: "Prasham Bhuta"
date: "08/04/2020"
header-includes:
    - \usepackage[sfdefault]{roboto}
    - \usepackage[T1]{fontenc}
output:
  html_document: default
  sans_font: default
  latex_engine: xelatex
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generating a plot that describes the US murder rate for all states based on the region.


The dataset is a part of the learning at Harvardx's Data Science Course.

### Loading required packages

```{r, warning=FALSE, message=FALSE}
library(dslabs)
library(tidyverse)
library(ggthemes)
library(ggrepel)
```

### Create a plot object with required data.
```{r}
p <- murders %>% ggplot(aes(population / 10^6, total, label=abb), height = 6, width = 16)
```

### Adding text geometry
```{r}
p <- p + geom_text_repel(nudge_x = 0.04, nudge_y = 0.04)
```

### Changing axis scale to be logarithmic
```{r}
p <- p + scale_x_log10() + scale_y_log10()
```

### Adding labels & titles
```{r}
p <- p + xlab("Population in millions (log)") +
     ylab("Total murders (log)") +
     ggtitle("US murders across states filtered by region")
```

## Calculate the average murder rate

This value will be plotted as line across the plot
```{r}
avg_rate <- murders %>% summarise(rate = sum(total) / (sum(population)) * 10^6) %>% .$rate
```

### Adding abline for the `avg_rate`
```{r}
p <- p + geom_abline(intercept = log10(avg_rate), color="orange", lty =2, size =1)
```

### Adding points as state with region as filter
```{r}
p <- p + geom_point(aes(col = region), size = 3)
```

### Renaming the label 'region`
```{r}
p <- p + scale_color_discrete(name = "Regions in US")
```

### Adding theme `economist` to our plot and improving font sizes
```{r}
p <- p + theme_economist() + theme(text = element_text(size = 16))
```

## The desired plot
```{r, echo=FALSE, fig.align='center', fig.width=10}
print(p)
```